---
title: "Practical Machine Learning Prediction Assignment"
author: "Anand Valmiki"
date: "Saturday, February 21, 2015"
output: html_document
---

### Housekeeping

```{r setup_env, echo=FALSE, warning=FALSE, message=FALSE}
WD <- c("C:/Users/20537710/My Documents/Coursera/Practical Machine Learning")
setwd(WD)

library(ggplot2)
library(caret)
library(randomForest)
```
```{r download, cache=TRUE}
fileUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
pmlTraining <- data.frame(read.csv(fileUrl, na.strings = c("", "NA"), header = TRUE))
```

### Data Cleanup

```{r misc-1}
dim(pmlTraining)
```
We find and remove columns with a large number NA's

```{r misc-2, echo=FALSE}
sapply(pmlTraining, function(x) sum(is.na(x)))
```

Using the above R fiunction we see that there are lots of variables which have a large umber of NA's to be specific we see that many rows have exactly 9216 NA's.
From reading the Forums I came to  know that the new_window column is yes, the corresponding row has summary statistics. This row of data is n We remove this row and columns which have all NA's.

```{r cleanup, echo=FALSE}
pmlTraining <- pmlTraining[! pmlTraining$new_window %in% c("yes") ,]
# Drops all columns with NA's
pmlTraining <- Filter(function(x)!all(is.na(x)), pmlTraining)

pmlTraining$magnet_dumbbell_z <- as.integer(pmlTraining$magnet_dumbbell_z)
pmlTraining$magnet_forearm_y <- as.integer(pmlTraining$magnet_forearm_y)
pmlTraining$magnet_forearm_z  <- as.integer(pmlTraining$magnet_forearm_z)

pmlTraining$raw_timestamp_part_2 <- NULL
pmlTraining$raw_timestamp_part_1 <- NULL
pmlTraining$cvtd_timestamp <- NULL
pmlTraining$new_window <- NULL
pmlTraining$num_window <- NULL
pmlTraining$X <- NULL
pmlTraining$user_name <- NULL
```

```{r train, echo=FALSE}
# Creating Training and Test data
inTrain <- createDataPartition(y = pmlTraining$classe, p = 0.75, list = FALSE)
training <- pmlTraining[inTrain ,]
testing <- pmlTraining[-inTrain ,]
#
rf <- randomForest(classe ~ ., data = training, ntree=100)
```
```{r results, echo=TRUE}
testing$rf <- predict(rf, testing)
table(testing$classe, testing$rf)
confusionMatrix(training$classe,predict(rf,training))
```
